Unbiased CVM Algorithm

### 2024-12-20 cvm_unbiased/main.tex
Minor improvements.

More practical versions of the previous lemma:
\begin{lemma}\label{le:neg_cor_pos}
Let $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$ be concave then for all 
$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
\[
  \expectation_{\Omega}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
\]
\end{lemma}
\begin{proof}
Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := h(\tau p^{-1})$. We just need to check Conditions 1/2, Indeed
\[
  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) = (1-\alpha) h(0) + \alpha h(x^{-1}) \leq h(\alpha x^{-1}) = \phi(x/\alpha,1)
\]
for $0 < \alpha < 1$ and $0< x \leq 1$, and
\[
  \phi(x,0) = h(0) \leq h(0) = \phi(y,0)
\]
for $0 < x < y \leq 1$.
\end{proof}

Unfortunately the above function is only applicable for monotone increasing functions, because a monotone decreasing function
will eventually be negative. The following modified version is for that case:

\begin{lemma}\label{le:neg_cor_neg}
Let $q \leq 1$ and $h : \mathbb [0,q^{-1}] \rightarrow \mathbb R_{\geq 0}$ be concave then for all 
$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
\[
  \expectation_{\Omega}\left[ \prod_{s \in S} I(p > q) h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
\]
\end{lemma}
\begin{proof}
Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just need to check Conditions 1/2, Indeed
\begin{eqnarray*}
  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) & = & (1-\alpha) I(x>q) h(0) + \alpha I(x > q) h(x^{-1}) \\
   & \leq & I(x > q) h(\alpha x^{-1}) \leq I(x > q \alpha) h(\alpha x^{-1}) = \phi(x/\alpha,1)
\end{eqnarray*}
for $0 < \alpha < 1$ and $0< x \leq 1$, where we used that $x > q$ implies $x > q \alpha$. And
\[
  \phi(x,0) = I(x>q) h(0) \leq I(y > q) h(0) = \phi(y,0)
\]
for $0 < x < y \leq 1$, where we used that $x > q$ implies $y > q$.
\end{proof}

@@

We define $q := n/(4|A|)$ - notice that $q \leq \frac{1}{4}$. Moreover we'll introduce the notation $\tilde{p} := \max(p,q)$.
Basically $\tilde{p}$ is a clamped version of $p$, and in particular, $\tilde{p}^{-1}$ is bounded from above by $q^{-1}$.

The following is a helper lemma:
\begin{lemma}
\label{le:helper}
For any $\Omega \in \{\Omega_0,\ldots,\Omega_l\} \cup \{\Omega'_1,\ldots,\Omega'_l\}$ and $0 < \eps \leq 1$:
\[
\prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A| \right) \leq \exp\left(-\frac{n}{12} \eps^2\right)
\]
\end{lemma}
\begin{proof}
By assumption there exists a $k$ such that $\Omega \in \{\Omega_k, \Omega'_k\}$. Let $A' = A \cap \{a_1,\ldots,a_k\}$.

@@

To get a tail estimate, we use the Cram\'{e}r-Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A| \right) & \leq & \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A'| \right) \\
  & \underset{\textrm{Markov,}t>0}{\leq} & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
 & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\
 & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} h( p^{-1} I(s \in \chi) ) \right] \\
 & \underset{\textrm{Le.~\ref{le:neg_cor_pos}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
 & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
\end{eqnarray*}}
So we just need to show that (using $|A|=\frac{n}{4q}$):
\[
  \ln(h(1))-t (1+\eps) \leq \frac{-q \eps^2}{3}
\]
The latter can be established by analyzing the function $f(q,\eps) = -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3}$.
For which it is easy to check $f(q,0) = 0$ and the derivative with respect to $\eps$ is positive in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
$f(q,\eps) \geq 0$.
\end{proof}

@@

Then
\begin{eqnarray*}
  \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( \tilde{p}^{-1} |\chi| \geq f^{-j} n ) \\
    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| ) \\
    & \underset{\textrm{Le.~\ref{le:helper}}}{\leq} & \exp( - n/12)
\end{eqnarray*}

@@

With these definitions we again follow the Cram\'{e}r-Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  L & \underset{t<0}{=} & \prob_{\Omega_l} \left( \exp(t p^{-1} |\chi|) \geq exp(t (1-\eps)|A|) \wedge p \geq q \right) \\
    & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq exp(t (1-\eps)|A|) \wedge p > q \right) \\
    & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
    & = & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\
    & \leq & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
    & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
    & = & \exp( \ln(h(1)) -t (1-\eps) )^{|A|}
\end{eqnarray*}
}

>>>>

More practical versions of the previous lemma:
\begin{lemma}\label{le:neg_cor_neg}
Let $q \leq 1$ and $h : \mathbb [0,q^{-1}] \rightarrow \mathbb R_{\geq 0}$ be concave then for all 
$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
\[
  \expectation_{\Omega}\left[ \prod_{s \in S} I(p > q) h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
\]
\end{lemma}
\begin{proof}
Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just need to check Conditions 1/2, Indeed
\begin{eqnarray*}
  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) & = & (1-\alpha) I(x>q) h(0) + \alpha I(x > q) h(x^{-1}) \\
   & \leq & I(x > q) h(\alpha x^{-1}) \leq I(x > q \alpha) h(\alpha x^{-1}) = \phi(x/\alpha,1)
\end{eqnarray*}
for $0 < \alpha < 1$ and $0< x \leq 1$, where we used that $x > q$ implies $x > q \alpha$. And
\[
  \phi(x,0) = I(x>q) h(0) \leq I(y > q) h(0) = \phi(y,0)
\]
for $0 < x < y \leq 1$, where we used that $x > q$ implies $y > q$.
\end{proof}

@@

We define $q := n/(4|A|)$ - notice that $q \leq \frac{1}{4}$.

The following is a helper lemma:
\begin{lemma}
\label{le:helper}
For any $\Omega \in \{\Omega_0,\ldots,\Omega_l\} \cup \{\Omega'_1,\ldots,\Omega'_l\}$ and $0 < \eps \leq 1$:
\[
\prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \right \wedge p \geq q) \leq \exp\left(-\frac{n}{12} \eps^2\right)
\]
\end{lemma}
\begin{proof}
By assumption there exists a $k$ such that $\Omega \in \{\Omega_k, \Omega'_k\}$. Let $A' = A \cap \{a_1,\ldots,a_k\}$.

@@

To get a tail estimate, we use the Cram\'{e}r-Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \right \wedge p \geq q) & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A'|) \right \wedge p \geq q) \\
  & \leq & \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
  & \underset{\textrm{Markov}}{\leq} & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
 & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
 & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
 & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
 & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
\end{eqnarray*}}
So we just need to show that (using $|A|=\frac{n}{4q}$):
\[
  \ln(h(1))-t (1+\eps) \leq \frac{-q \eps^2}{3}
\]
The latter can be established by analyzing the function $f(q,\eps) = -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3}$.
For which it is easy to check $f(q,0) = 0$ and the derivative with respect to $\eps$ is positive in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
$f(q,\eps) \geq 0$.
\end{proof}

@@

Then
\begin{eqnarray*}
  \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( p^{-1} |\chi| \geq f^{-j} n \wedge p \geq q) \\
    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| \wegde p \geq q ) \\
    & \underset{\textrm{Le.~\ref{le:helper}}}{\leq} & \exp(- n/12)
\end{eqnarray*}

@@

With these definitions we again follow the Cram\'{e}r-Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  L & \underset{t<0}{=} & \prob_{\Omega_l} \left( \exp(t p^{-1} |\chi|) \geq exp(t (1-\eps)|A|) \wedge p \geq q \right) \\
    & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq exp(t (1-\eps)|A|) \wedge p > q \right) \\
    & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
    & = & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
    & \leq & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
    & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
    & = & \exp( \ln(h(1)) -t (1-\eps) )^{|A|}
\end{eqnarray*}
}
